Supplemental Material: Multidimensional analysis and detection of informative features in human brain white matter

### 2021-06-09 reports/manuscript/paper_changes.tex
Accept all changes

\replaced{\mbox{\cite{Jones2005-yq, yeatman2012tract, colby2012, ODonnell2009-uu}},}{\mbox{\cite{yeatman2012tract, colby2012, ODonnell2009-uu}},}
\added{hereafter referred to as tract profiling}.

@@

The present work aims to balance predictive accuracy with descriptive power
\cite{Murdoch2019-ax, Breiman2001-uz} by capitalizing on all of the available
data across all bundles, while also retaining and elucidating spatial
information about the locations that are most informative for discriminative
performance.
\added{Unlike many previous analyses of white matter (e.g., using TBSS \cite{smith2006tract} and connectometry \cite{yeh2016connectometry}), which focus on classical inference about the differences between groups or individuals in terms of their white matter properties, the focus of the present work is on the combination of features that facilitate accurate prediction of the individual differences in a particular phenotype}.
\added{For example, an accurate prediction of which individuals in a group have a particular disease, or a prediction of their age with small error. The distinct goals of \emph{inference} and \emph{prediction} are not necessarily incongruent, but can sometimes be in tension \cite{Bzdok2020-sd}. This distinction is also similar to the distinction between ``encoding'' and ``decoding'' used in the functional MRI literature \cite{Naselaris2011-ss}. In the present work}, we predict the phenotypical variance in a group of subjects,or classify group membership, based on a linear combination of the features estimated with \replaced{tract profiling}{tractometry}.

@@

Using this approach, we first need to deal with the large and
asymmetric dimensionality of the data: \replaced{tract profile}{tractometry}
data usually has
many more features (i.e., number of measurements per individual) than
samples (number of subjects), which makes inferences from the data
about phenotypical differences between individuals ill-posed. This
regime is the target of several statistical learning techniques, and
is often solved by various forms of regularization.

% For example, Tikhonov regularization shrinks the solution such that the sum of
% squared contributions from the individual features are minimized
% \cite{Hoerl2000-ij}. Another solution to the problem is provided by
% the
The Lasso algorithm minimizes the sum of the absolute values of
contributions of each feature \cite{Tibshirani1996-qs}. This
tends to shrink to zero the contributions of many of the features,
providing results that are both accurate and interpretable. When
additional structure is available in the organization of the data,
regularization algorithms can take advantage of this structure. For
example, if the features lend themselves to a natural division into
different groups, the Group Lasso (GL) can be used to select groups
of features, rather than individual features \cite{Yuan2006-ky}.
The Sparse Group Lasso (SGL) elaborates on this idea by providing
control both of group sparsity, as well as overall sparsity of the
solutions \cite{simon2013sgl}. Because the features measured with
tractomery lend themselves to grouping based on the tracts from which
each measurement is taken, GL and SGL provide useful tools
for linear model fitting in problems of this form. Here we, first,
develop an implementation of SGL that is well suited to the analysis of
\replaced{tract profile}{tractometry} data and, second, demonstrate the power and flexibility of
this approach by applying it to both classification
and continuous prediction problems. Our \deleted{tractometry} data flow is
represented in \cref{fig:methods} and explained in further detail in
the methods section.

\added{One more approach to multivariate analysis of neuroimaging data is to perform inference or prediction in a transformed feature space rather then the original diffusion metric feature space. For example, in Lasso PCR, diffusion features can be projected onto a principal components (PC) basis for use in a sparsity constrained model \mbox{\cite{powell2018local, rasero2021integrating}}. While these approaches achieve high predictive performance and may even yield biologically interpretable principal components \mbox{\cite{chamberland2019dimensionality}}, they neglect anatomical grouping information. We view these approaches as complementary to the SGL-based approach: PCR based approaches seeks to model brain-phenotype relationships using the most parsimonious representation of variance in the diffusion measures. In contrast, our SGL-based approach seeks to establish whether prior knowledge of anatomical grouping improves modeling of brain-phenotype relationships. Motivated by this distinction, we also introduce a union of the two approaches called PCR-SGL, in which each group of features is independently transformed into its PC basis, thereby retaining the anatomical grouping information of the original feature space. We demonstrate circumstances in which this approach both helps and hinders predictive performance.}

\begin{figure}[b!]
    % \vspace{0.25cm}
    % \rlap{
    %     \hspace{-0.25em}
    %     \includegraphics[width=0.98\columnwidth]{methods-brains.pdf}
    % }
    % \vspace{-8.4cm}
    % \includegraphics[width=\columnwidth]{method-quad.pdf}
    {\phantomsubcaption\label{fig:methods:tractogram}}
    {\phantomsubcaption\label{fig:methods:cst}}
    {\phantomsubcaption\label{fig:methods:tract-profile:cst}}
    {\phantomsubcaption\label{fig:methods:tract-profile:fa}}
    {\phantomsubcaption\label{fig:methods:group-structure}}
    \caption{\label{fig:methods}
    {\bf Figure 1: Tractometry data flow}
    \textbf{(a)} Whole brain tractography generates streamlines approximating
        the trajectories of white matter connections.
        \textbf{(b)} Tractometry classifies these streamlines into anatomical bundles.
        In this case, we show the left corticospinal tract (CSTL)
        \added{and the left arcuate fasciculus (ARCL)}
        over a mid-saggital anatomical slice.
        \deleted{\textbf{(c)}} \replaced{Tract profiling}{Tractometry} further extracts bundle profiles,
        quantifications of various diffusion metrics along the length of the
        fiber bundle. Here, we show one subject's fractional anisotropy (FA)
        profile for \added{\textbf{(c)}} the CSTL \added{and \textbf{(d)} the ARCL}.
        \textbf{(d)} the phenotypical target data and \replaced{tract profile}{tractometric}
        features can be organized into a linear model, $\hat{y} = \mathbf{X}
        \hat{\beta}$. The feature matrix $\mathbf{X}$ is color-coded
        to reveal a natural group structure: the left (orange) group
        contains $k$ features from the CSTL, the middle (green) group
        contains $k$ features from the left cingulum cingulate (CGCL),
        and the right (blue) group
        contains $k$ features from the \replaced{ARCL}{left arcuate (ARCL)}.
        The coefficients in $\hat{\beta}$ follow the same natural grouping. \added{Panels (a) and (b) are adapted from \url{https://figshare.com/articles/figure/example_tractography-segmentation/14485350}, and reproduced under the CC-BY license (\url{https://creativecommons.org/licenses/by/4.0/})}
    }
\end{figure}

@@

We developed a method for analyzing dMRI \replaced{tract profile}{tractometry} data that uses the Sparse
Group Lasso (SGL) to select features that are sparse both at the group (bundle) level, as well as overall.
We demonstrate the use of this method on four different datasets in
both a classification setting and a regression setting.

@@

Using data from a previous study of patients with amyotrophic lateral sclerosis (ALS)~\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on \emph{a priori}
selection of features only within the CST bundle-of-interest.
SGL delivers improved predictive performance, with a cross-validated accuracy
of {\alsAccuracy}\% and an area under the receiver operating characteristic
curve (ROC AUC) of {\alsRocAuc}, without the need for \emph{a priori} feature
engineering.
\added{We also predicted ALS diagnosis using the PCR-SGL, wherein each group of features is independently transformed to its PC basis and an SGL model is fit to the transformed features. This model achieves \protect{\alsAccuracyGpca}\% accuracy and an ROC AUC of \protect{\alsRocAucGpca}}
(\cref{fig:class-results:probs}). In addition to this classification
performance, both SGL \added{and PCR-SGL} also \replaced{identify}{identifies}
the white matter tracts most important for ALS classification.
\added{\Cref{fig:class-results:coefs3d} shows that PCR-SGL identified as potential disease biomarkers the diffusion measures in the CST from the cerebral peduncle to the corona radiata, agreeing with the previous study from which these data were extracted \mbox{\cite{sarica2017corticospinal}}.}
The relative importance of white matter
features is captured in the $\beta$ coefficients from \cref{eq:sgl}.
\Cref{fig:class-results:tract-profiles} depicts these coefficients along the
right CST, plotted over the FA values for the control and ALS subject groups \added{(see Supporting Figures S1 and S2 for tract profiles of all 18 tracts in these groups)}.
We find that SGL \replaced{and PCR-SGL select}{selects} FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings \cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in
the previous study from which we obtained the data
\cite{sarica2017corticospinal}.

To assess the added value of incorporating anatomical knowledge into our
model, we compared our classification results to those obtained from
\replaced{four other models: a pure lasso model, an elastic net model, a bundle-mean lasso (i.e. a lasso model trained only on the mean metric values from each bundle), and a principal components regression lasso (Lasso PCR) model \cite{powell2018local, rasero2021integrating}}{a lasso model}.
\replaced{These models achieved accuracies of 76\%, 76\%, 79.5\%, and 71.5\% respectively and ROC AUC values of 0.76, 0.75, 0.82, and 0.71, respectively}{The pure lasso model achieved 76\% accuracy and an ROC AUC of 0.76}
using the same cross-validation strategy used to assess the SGL model.
This difference in performance justifies the additional complexity of the SGL
over simpler sparsity-inducing regression strategies.
\added{The relatively poor performance of Lasso PCR suggests that features with small variance are nonetheless relevant for predicting ALS diagnosis. Or equivalently, that group differences between ALS diagnoses are not the dominant source of variance in the diffusion metrics.}

The $\beta$ coefficients exhibit high bundle level sparsity; only some
bundles are important, which can be be confirmed by observing the value of
$\alpha$, the regularization hyperparameter that controls the mixture of the
Group Lasso and lasso penalties, selected through nested cross-validation
(see \cref{eq:sgl}).
If $\alpha$ is closer to zero, it indicates that the phenotype in question
preferentially correlates with only a few groups of covariates. For the ALS
dataset, \added{the SGL model has} $\alpha = $ \alsLRatio
~\added{and the PCR-SGL model has \protect$\alpha = $ \protect\alsLRatioGpca},
confirming that the white matter correlates of ALS reside mostly in one
bundle, namely the CST.

@@

\begin{figure}[b!]
    % \vspace{3.0cm}
    % \rlap{
    %     \includegraphics[width=0.33\textwidth]{als_gpca_coefs.pdf}
    % }
    % \vspace{-7.0cm}
    % \includegraphics[width=\textwidth]{als_gpca_results.pdf}
    {\phantomsubcaption\label{fig:class-results:probs}}
    {\phantomsubcaption\label{fig:class-results:coefs3d}}
    {\phantomsubcaption\label{fig:class-results:tract-profiles}}
    {\phantomsubcaption\label{fig:class-results:false-positives}}
    {\phantomsubcaption\label{fig:class-results:false-negatives}}
    \caption{
    {\bf Figure 2: \added{PCR-}SGL accurately and interpretably predicts ALS diagnosis.}
    {\bf (a)} Classification probabilities for ALS diagnosis, with
    controls on the left, patients on the right, predicted controls in
    blue, and predicted patients in orange. That is, orange dots on the left
    represent false positives, while blue dots on the right represent
    false negatives. We achieve {\protect\alsAccuracy}\% accuracy with an
    ROC AUC of {\protect\alsRocAuc}.
    {\bf (b)} \added{PCR-}SGL coefficients are presented on the core fibers of major
    fiber bundles. They exhibit high group sparsity and are concentrated
    in the FA of the \replaced{corticospinal tract (CST)}{CST}.
    The brain is oriented with the right hemisphere
    in the foreground and anterior to the right of the page. The CSTL, CSTR,
    \replaced{callosum forceps anterior (CFA), left arcuate (ARCL), and right arcuate (ARCR)}{CFA}
    bundles are indicated for orientation.
    {\bf (c)} \added{PCR-}SGL identifies three portions of the CST as important,
    where $\hat{\beta}$ (dashed line, right axis) has large values. These
    are centered around nodes 30, 65, and 90, corresponding to locations
    of substantial differences in FA between the ALS and control groups
    (shaded areas indicates standard error of the mean).
    {\bf (d)} Bundle profiles for false positive classifications. Line
    colors correspond to the marker edge color in the top left plot.
    These individuals have reduced FA in the CST portions which SGL
    identified as important. Their misclassification is coherent with the
    feature importance and the group differences in FA.
    {\bf (e)} Individual bundle profiles for false negative
    classifications. These individuals have bundle profiles which
    oscillate between the group means.
    \label{fig:class-results}
    }
\end{figure}

@@

To test the performance of SGL \deleted{with tractometry data} in a continuous
regression task, we focus here on the prediction of biological age in three
datasets named WH, HBN, and Cam-CAN (see the Methods section for a description of each dataset). Prediction of ``brain
age'' is a commonly undertaken task in neuroimaging machine learning, in part
because these predictions, and deviations therefrom, may be diagnostic of
overall brain health (for a review, see Cole et al.~\cite{Cole2019-rz}).
However, as Nelson et al.~\cite{nelson2019biomarkers} have observed, aging
biomarkers are subject to unique challenges and tend to be noncausative. Our
interest in aging here relies on its utility as a methodological benchmark.
Biological age operates on a natural scale, with meaningful and easily
understood units, and it's popularity as a machine learning target makes it
valuable for comparisons with other studies.

The WH \cite{yeatman2014lifespan}, HBN \cite{alexander2017open}, and Cam-CAN
\cite{shafto2014cambridge,taylor2017cambridge} datasets used here contain
data from 76, 1651, and 640 subjects, respectively, ranging from 6-50, 5-21,
and 18-88 years of age, respectively. In each case, biological age was used
as the predicted variable $y$. SGL was fit to the \replaced{tract profile}{tractometry-extracted}
features FA and mean diffusivity (MD) in 18 major brain tracts, with diffusion metrics extracted
from diffusion tensor imaging (DTI) for the WH dataset and diffusion kurtosis
imaging (DKI) \cite{jensen2005diffusion} for the HBN and Cam-CAN datasets \added{see Supporting Figures S3 to S8 for the full tract profile information in all tracts/datasets}.

To evaluate the fit of the model, we used a nested cross-validation
procedure. In this procedure, batches of subjects are held out. For each
batch (or fold), the model is fully fit without this data. Then, once the
parameters are fixed, the model is applied to predict the ages of held out
subjects based on the linear coeffiecients. This scheme automatically finds
the right level of sparsity and fits the coefficients to the ill-posed linear
model, while guarding against overfitting. SGL accurately predicts the age of
the subjects in this procedure, with a median absolute error of {\whMae},
\added{{\hbnMae}}, and {\camcanMae} years for the WH, HBN, and Cam-CAN datasets,
respectively and coefficients of determination $R^2 = $ {\whRsq} , \added{{\hbnRsq}}, and {\camcanRsq},
respectively (see \cref{fig:age-results}, top panels). The
predictions for Cam-CAN are competitive with a recent state of the art
prediction \cite{mcpherson2020single}, which used streamline density to
estimate the brain's structural connectivity and achieved $R^2 = 0.63$. The
median absolute errors are also lower than the results of a recent study that
predicted age in a large sample that included the Cam-CAN data, and was based
on diffusion MRI features \cite{Richard2018-ux}.

@@

As with ALS classification, we also compared SGL performance with results
obtained using
\replaced{the pure lasso, elastic net, bundle-mean lasso, and Lasso PCR}{only the lasso penalty}.
The pure lasso models achieved $R^2 = 0.47$, $0.54$, and $0.70$ for the WH, HBN, and Cam-CAN datasets respectively using the same cross-validation strategy used to assess the SGL.
\added{The elastic net models achieved $R^2 = 0.46$, $0.59$, and $0.73$ for the WH, HBN, and Cam-CAN datasets, respectively. The bundle-mean lasso models achieved $R^2 = 0.33$, $0.22$, and $0.55$, respectively. And the Lasso PCR models achieved $R^2 = 0.54$, $0.57$, $0.74$, respectively.}
The SGL models'
performance improvement over lasso is more modest than in the ALS
classification case, which accords with the selected values of the $\alpha$
hyperparameter. These SGL models were more lasso-like in their sparsity
penalties so they are more lasso-like in their predictive performance.

\added{Model performance across all four datasets and all six model types is summarized}
\added{in \cref{fig:model-performance}. In contrast to the ALS classification case, the PCR Lasso performs}
\added{competitively in age regression, suggesting that aging is a significant source of global variance}
\added{in the tract profiles. On the other hand, the PCR-SGL models destroy cross-bundle covariance information}
\added{in their initial group-wise PC projection step (see \cref{eq:svd}), limiting performance}
\added{for globally distributed phenomena. Conversely, the SGL models are able to adapt to this regime.}

>>>>

\cite{Jones2005-yq, yeatman2012tract, colby2012, ODonnell2009-uu},
hereafter referred to as tract profiling.

@@

The present work aims to balance predictive accuracy with descriptive power
\cite{Murdoch2019-ax, Breiman2001-uz} by capitalizing on all of the available
data across all bundles, while also retaining and elucidating spatial
information about the locations that are most informative for discriminative
performance.
Unlike many previous analyses of white matter (e.g., using TBSS \cite{smith2006tract} and connectometry \cite{yeh2016connectometry}), which focus on classical inference about the differences between groups or individuals in terms of their white matter properties, the focus of the present work is on the combination of features that facilitate accurate prediction of the individual differences in a particular phenotype.
For example, an accurate prediction of which individuals in a group have a particular disease, or a prediction of their age with small error. The distinct goals of \emph{inference} and \emph{prediction} are not necessarily incongruent, but can sometimes be in tension \cite{Bzdok2020-sd}. This distinction is also similar to the distinction between ``encoding'' and ``decoding'' used in the functional MRI literature \cite{Naselaris2011-ss}. In the present work, we predict the phenotypical variance in a group of subjects,or classify group membership, based on a linear combination of the features estimated with tract profiling.

@@

Using this approach, we first need to deal with the large and
asymmetric dimensionality of the data: tract profile
data usually has
many more features (i.e., number of measurements per individual) than
samples (number of subjects), which makes inferences from the data
about phenotypical differences between individuals ill-posed. This
regime is the target of several statistical learning techniques, and
is often solved by various forms of regularization.

% For example, Tikhonov regularization shrinks the solution such that the sum of
% squared contributions from the individual features are minimized
% \cite{Hoerl2000-ij}. Another solution to the problem is provided by
% the
The Lasso algorithm minimizes the sum of the absolute values of
contributions of each feature \cite{Tibshirani1996-qs}. This
tends to shrink to zero the contributions of many of the features,
providing results that are both accurate and interpretable. When
additional structure is available in the organization of the data,
regularization algorithms can take advantage of this structure. For
example, if the features lend themselves to a natural division into
different groups, the Group Lasso (GL) can be used to select groups
of features, rather than individual features \cite{Yuan2006-ky}.
The Sparse Group Lasso (SGL) elaborates on this idea by providing
control both of group sparsity, as well as overall sparsity of the
solutions \cite{simon2013sgl}. Because the features measured with
tractomery lend themselves to grouping based on the tracts from which
each measurement is taken, GL and SGL provide useful tools
for linear model fitting in problems of this form. Here we, first,
develop an implementation of SGL that is well suited to the analysis of
tract profile data and, second, demonstrate the power and flexibility of
this approach by applying it to both classification
and continuous prediction problems. Our data flow is
represented in \cref{fig:methods} and explained in further detail in
the methods section.

One more approach to multivariate analysis of neuroimaging data is to perform inference or prediction in a transformed feature space rather then the original diffusion metric feature space. For example, in Lasso PCR, diffusion features can be projected onto a principal components (PC) basis for use in a sparsity constrained model \cite{powell2018local, rasero2021integrating}. While these approaches achieve high predictive performance and may even yield biologically interpretable principal components \cite{chamberland2019dimensionality}, they neglect anatomical grouping information. We view these approaches as complementary to the SGL-based approach: PCR based approaches seeks to model brain-phenotype relationships using the most parsimonious representation of variance in the diffusion measures. In contrast, our SGL-based approach seeks to establish whether prior knowledge of anatomical grouping improves modeling of brain-phenotype relationships. Motivated by this distinction, we also introduce a union of the two approaches called PCR-SGL, in which each group of features is independently transformed into its PC basis, thereby retaining the anatomical grouping information of the original feature space. We demonstrate circumstances in which this approach both helps and hinders predictive performance.

\begin{figure}[b!]
    % \vspace{0.25cm}
    % \rlap{
    %     \hspace{-0.25em}
    %     \includegraphics[width=0.98\columnwidth]{methods-brains.pdf}
    % }
    % \vspace{-8.4cm}
    % \includegraphics[width=\columnwidth]{method-quad.pdf}
    {\phantomsubcaption\label{fig:methods:tractogram}}
    {\phantomsubcaption\label{fig:methods:cst}}
    {\phantomsubcaption\label{fig:methods:tract-profile:cst}}
    {\phantomsubcaption\label{fig:methods:tract-profile:fa}}
    {\phantomsubcaption\label{fig:methods:group-structure}}
    \caption{\label{fig:methods}
    {\bf Figure 1: Tractometry data flow}
    \textbf{(a)} Whole brain tractography generates streamlines approximating
        the trajectories of white matter connections.
        \textbf{(b)} Tractometry classifies these streamlines into anatomical bundles.
        In this case, we show the left corticospinal tract (CSTL)
        and the left arcuate fasciculus (ARCL)
        over a mid-saggital anatomical slice.
        Tract profiling further extracts bundle profiles,
        quantifications of various diffusion metrics along the length of the
        fiber bundle. Here, we show one subject's fractional anisotropy (FA)
        profile for \textbf{(c)} the CSTL and \textbf{(d)} the ARCL.
        \textbf{(d)} the phenotypical target data and tract profile
        features can be organized into a linear model, $\hat{y} = \mathbf{X}
        \hat{\beta}$. The feature matrix $\mathbf{X}$ is color-coded
        to reveal a natural group structure: the left (orange) group
        contains $k$ features from the CSTL, the middle (green) group
        contains $k$ features from the left cingulum cingulate (CGCL),
        and the right (blue) group
        contains $k$ features from the ARCL.
        The coefficients in $\hat{\beta}$ follow the same natural grouping. Panels (a) and (b) are adapted from \url{https://figshare.com/articles/figure/example_tractography-segmentation/14485350}, and reproduced under the CC-BY license (\url{https://creativecommons.org/licenses/by/4.0/})
    }
\end{figure}

@@

We developed a method for analyzing dMRI tract profile data that uses the Sparse
Group Lasso (SGL) to select features that are sparse both at the group (bundle) level, as well as overall.
We demonstrate the use of this method on four different datasets in
both a classification setting and a regression setting.

@@

Using data from a previous study of patients with amyotrophic lateral sclerosis (ALS)~\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on \emph{a priori}
selection of features only within the CST bundle-of-interest.
SGL delivers improved predictive performance, with a cross-validated accuracy
of {\alsAccuracy}\% and an area under the receiver operating characteristic
curve (ROC AUC) of {\alsRocAuc}, without the need for \emph{a priori} feature
engineering.
We also predicted ALS diagnosis using the PCR-SGL, wherein each group of features is independently transformed to its PC basis and an SGL model is fit to the transformed features. This model achieves \protect{\alsAccuracyGpca}\% accuracy and an ROC AUC of \protect{\alsRocAucGpca}
(\cref{fig:class-results:probs}). In addition to this classification
performance, both SGL and PCR-SGL also identify
the white matter tracts most important for ALS classification.
\Cref{fig:class-results:coefs3d} shows that PCR-SGL identified as potential disease biomarkers the diffusion measures in the CST from the cerebral peduncle to the corona radiata, agreeing with the previous study from which these data were extracted \cite{sarica2017corticospinal}.
The relative importance of white matter
features is captured in the $\beta$ coefficients from \cref{eq:sgl}.
\Cref{fig:class-results:tract-profiles} depicts these coefficients along the
right CST, plotted over the FA values for the control and ALS subject groups (see Supporting Figures S1 and S2 for tract profiles of all 18 tracts in these groups).
We find that SGL and PCR-SGL select FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings \cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in
the previous study from which we obtained the data
\cite{sarica2017corticospinal}.

To assess the added value of incorporating anatomical knowledge into our
model, we compared our classification results to those obtained from
four other models: a pure lasso model, an elastic net model, a bundle-mean lasso (i.e. a lasso model trained only on the mean metric values from each bundle), and a principal components regression lasso (Lasso PCR) model \cite{powell2018local, rasero2021integrating}.
These models achieved accuracies of 76\%, 76\%, 79.5\%, and 71.5\% respectively and ROC AUC values of 0.76, 0.75, 0.82, and 0.71, respectively
using the same cross-validation strategy used to assess the SGL model.
This difference in performance justifies the additional complexity of the SGL
over simpler sparsity-inducing regression strategies.
The relatively poor performance of Lasso PCR suggests that features with small variance are nonetheless relevant for predicting ALS diagnosis. Or equivalently, that group differences between ALS diagnoses are not the dominant source of variance in the diffusion metrics.

The $\beta$ coefficients exhibit high bundle level sparsity; only some
bundles are important, which can be be confirmed by observing the value of
$\alpha$, the regularization hyperparameter that controls the mixture of the
Group Lasso and lasso penalties, selected through nested cross-validation
(see \cref{eq:sgl}).
If $\alpha$ is closer to zero, it indicates that the phenotype in question
preferentially correlates with only a few groups of covariates. For the ALS
dataset, the SGL model has $\alpha = $ \alsLRatio
~and the PCR-SGL model has \protect$\alpha = $ \protect\alsLRatioGpca,
confirming that the white matter correlates of ALS reside mostly in one
bundle, namely the CST.

@@

\begin{figure}[b!]
    % \vspace{3.0cm}
    % \rlap{
    %     \includegraphics[width=0.33\textwidth]{als_gpca_coefs.pdf}
    % }
    % \vspace{-7.0cm}
    % \includegraphics[width=\textwidth]{als_gpca_results.pdf}
    {\phantomsubcaption\label{fig:class-results:probs}}
    {\phantomsubcaption\label{fig:class-results:coefs3d}}
    {\phantomsubcaption\label{fig:class-results:tract-profiles}}
    {\phantomsubcaption\label{fig:class-results:false-positives}}
    {\phantomsubcaption\label{fig:class-results:false-negatives}}
    \caption{
    {\bf Figure 2: PCR-SGL accurately and interpretably predicts ALS diagnosis.}
    {\bf (a)} Classification probabilities for ALS diagnosis, with
    controls on the left, patients on the right, predicted controls in
    blue, and predicted patients in orange. That is, orange dots on the left
    represent false positives, while blue dots on the right represent
    false negatives. We achieve {\protect\alsAccuracy}\% accuracy with an
    ROC AUC of {\protect\alsRocAuc}.
    {\bf (b)} PCR-SGL coefficients are presented on the core fibers of major
    fiber bundles. They exhibit high group sparsity and are concentrated
    in the FA of the corticospinal tract (CST).
    The brain is oriented with the right hemisphere
    in the foreground and anterior to the right of the page. The CSTL, CSTR,
    callosum forceps anterior (CFA), left arcuate (ARCL), and right arcuate (ARCR)
    bundles are indicated for orientation.
    {\bf (c)} PCR-SGL identifies three portions of the CST as important,
    where $\hat{\beta}$ (dashed line, right axis) has large values. These
    are centered around nodes 30, 65, and 90, corresponding to locations
    of substantial differences in FA between the ALS and control groups
    (shaded areas indicates standard error of the mean).
    {\bf (d)} Bundle profiles for false positive classifications. Line
    colors correspond to the marker edge color in the top left plot.
    These individuals have reduced FA in the CST portions which SGL
    identified as important. Their misclassification is coherent with the
    feature importance and the group differences in FA.
    {\bf (e)} Individual bundle profiles for false negative
    classifications. These individuals have bundle profiles which
    oscillate between the group means.
    \label{fig:class-results}
    }
\end{figure}

@@

To test the performance of SGL in a continuous
regression task, we focus here on the prediction of biological age in three
datasets named WH, HBN, and Cam-CAN (see the Methods section for a description of each dataset). Prediction of ``brain
age'' is a commonly undertaken task in neuroimaging machine learning, in part
because these predictions, and deviations therefrom, may be diagnostic of
overall brain health (for a review, see Cole et al.~\cite{Cole2019-rz}).
However, as Nelson et al.~\cite{nelson2019biomarkers} have observed, aging
biomarkers are subject to unique challenges and tend to be noncausative. Our
interest in aging here relies on its utility as a methodological benchmark.
Biological age operates on a natural scale, with meaningful and easily
understood units, and it's popularity as a machine learning target makes it
valuable for comparisons with other studies.

The WH \cite{yeatman2014lifespan}, HBN \cite{alexander2017open}, and Cam-CAN
\cite{shafto2014cambridge,taylor2017cambridge} datasets used here contain
data from 76, 1651, and 640 subjects, respectively, ranging from 6-50, 5-21,
and 18-88 years of age, respectively. In each case, biological age was used
as the predicted variable $y$. SGL was fit to the tract profile
features FA and mean diffusivity (MD) in 18 major brain tracts, with diffusion metrics extracted
from diffusion tensor imaging (DTI) for the WH dataset and diffusion kurtosis
imaging (DKI) \cite{jensen2005diffusion} for the HBN and Cam-CAN datasets see Supporting Figures S3 to S8 for the full tract profile information in all tracts/datasets.

To evaluate the fit of the model, we used a nested cross-validation
procedure. In this procedure, batches of subjects are held out. For each
batch (or fold), the model is fully fit without this data. Then, once the
parameters are fixed, the model is applied to predict the ages of held out
subjects based on the linear coeffiecients. This scheme automatically finds
the right level of sparsity and fits the coefficients to the ill-posed linear
model, while guarding against overfitting. SGL accurately predicts the age of
the subjects in this procedure, with a median absolute error of {\whMae},
{\hbnMae}, and {\camcanMae} years for the WH, HBN, and Cam-CAN datasets,
respectively and coefficients of determination $R^2 = $ {\whRsq} , {\hbnRsq}, and {\camcanRsq},
respectively (see \cref{fig:age-results}, top panels). The
predictions for Cam-CAN are competitive with a recent state of the art
prediction \cite{mcpherson2020single}, which used streamline density to
estimate the brain's structural connectivity and achieved $R^2 = 0.63$. The
median absolute errors are also lower than the results of a recent study that
predicted age in a large sample that included the Cam-CAN data, and was based
on diffusion MRI features \cite{Richard2018-ux}.

@@

As with ALS classification, we also compared SGL performance with results
obtained using
the pure lasso, elastic net, bundle-mean lasso, and Lasso PCR.
The pure lasso models achieved $R^2 = 0.47$, $0.54$, and $0.70$ for the WH, HBN, and Cam-CAN datasets respectively using the same cross-validation strategy used to assess the SGL.
The elastic net models achieved $R^2 = 0.46$, $0.59$, and $0.73$ for the WH, HBN, and Cam-CAN datasets, respectively. The bundle-mean lasso models achieved $R^2 = 0.33$, $0.22$, and $0.55$, respectively. And the Lasso PCR models achieved $R^2 = 0.54$, $0.57$, $0.74$, respectively.
The SGL models'
performance improvement over lasso is more modest than in the ALS
classification case, which accords with the selected values of the $\alpha$
hyperparameter. These SGL models were more lasso-like in their sparsity
penalties so they are more lasso-like in their predictive performance.

Model performance across all four datasets and all six model types is summarized
in \cref{fig:model-performance}. In contrast to the ALS classification case, the PCR Lasso performs
competitively in age regression, suggesting that aging is a significant source of global variance
in the tract profiles. On the other hand, the PCR-SGL models destroy cross-bundle covariance information
in their initial group-wise PC projection step (see \cref{eq:svd}), limiting performance
for globally distributed phenomena. Conversely, the SGL models are able to adapt to this regime.

\begin{figure}[b!]
    % \vspace{3.0cm}
    % \rlap{
    %     \hspace{0.775cm}
    %     \includegraphics[width=0.26\textwidth]{wh_coefs.pdf}
    %     \hspace{0.5em}
    %     \includegraphics[width=0.26\textwidth]{hbn_coefs.pdf}
    %     \hspace{0.5em}
    %     \includegraphics[width=0.26\textwidth]{camcan_coefs.pdf}
    % }
    % \vspace{-6.1cm}
    % \includegraphics[width=\textwidth]{regression_scatter.pdf}
    {\phantomsubcaption\label{fig:age-results:wh-scatter}}
    {\phantomsubcaption\label{fig:age-results:hbn-scatter}}
    {\phantomsubcaption\label{fig:age-results:cc-scatter}}
    {\phantomsubcaption\label{fig:age-results:wh-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:hbn-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:cc-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:wh-profile}}
    {\phantomsubcaption\label{fig:age-results:hbn-profile}}
    {\phantomsubcaption\label{fig:age-results:cc-profile}}
    \phantomcaption{%
    {\bf Figure 3: Predicting age with tractometry and SGL.}
    {\bf (top)} The predicted age vs. true age of each individual from the test
splits (i.e., when each subject's data was held out in fitting the model) for
the {\bf (a)} WH, {\bf (b)} HBN, and {\bf (c)} Cam-CAN datasets; an accurate
prediction falls close to the $y=x$ line (dashed). The mean absolute error (MAE)
and coefficient of determination $R^2$ are presented in the lower right of each
scatter plot. {\bf (middle)} Feature importance for predicting age from
tract profile in the {\bf (d)} WH, {\bf (e)} HBN, and
{\bf (f)} Cam-CAN datasets. The orientation of the brain is that same as in
\cref{fig:class-results:coefs3d}, however because the coefficients exhibit high
global sparsity (as opposed to group sparsity), we plot the mean of the absolute
value of $\hat{\beta}$ for each bundle on the core fiber. The global
distrubution of the $\hat{\beta}$ coefficients reflects the fact that aging is
not confined to a single white matter bundle. {\bf (bottom)} Age quintile bundle
profiles for the {\bf (g)} WH, {\bf (h)} HBN, and {\bf (i)} Cam-CAN datasets.
    \label{fig:age-results}
    }
\end{figure}

\begin{figure}[b!]
    % \includegraphics[width=\textwidth]{model_performance.pdf}
    \caption{%
    {\bf Figure 4: Model performance across all datasets.}
Each panel shows model performance measured on the test set for each
cross-validation split, with each black dot representing a split, box
plots representing the quartiles, and white diamonds representing the mean
performance. The $y$-scale varies in each subplot.
\textbf{(a)} Accuracy of test set predictions for the
ALS dataset. Because group differences
in ALS diagnosis are mostly confined to a single bundle, the group
structure-preserving methods, SGL and PCR-SGL, outperform the other
models.
The remaining frames show coefficient of determination, \protect$R^2$ in test sets
for the \textbf{(b)} WH, \textbf{(c)} HBN, and \textbf{(c)} Cam-CAN
datasets. Because aging affects the white matter globally, group structure-blind
methods like elastic net and PCR Lasso perform well. Nonetheless, the SGL
models show competitve predictive performance, adapting to a problem where
group structure is not as informative. PCR-SGL performs poorly in this regime
because its initial group-wise PC projection destroys between bundle covariance.
The bundle-mean lasso performs poorly, demonstrating the value of along-tract profiling.
    \label{fig:model-performance}
    }
\end{figure}
